### 2020-12-07 slides.Rmd
Add files via upload

---
title: "DSCI 445 - Utilizing Machine Learning for Algae Predictions"
author: "Dawson Eliasen, Boston Lee, Miles Austin"
date: "12/7/2020"
output:
  slidy_presentation: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(tmap)
library(tmaptools)
library(USAboundaries)
library(sf)
knitr::opts_chunk$set(echo = FALSE)
```

## Motivation

When looking at the amount of hand-collected (in situ) data, we see that in Colorado only has data for the major municipal water bodies like Horsetooth.
```{r, eval=F}
sampled_map <- st_read("sampled_map.csv", geometry_column='geometry')
states <- us_boundaries()
st_as_sf("sampled_map")
co_outline <- states %>%
  filter(name == 'Colorado') %>%
  st_transform(2163)
tmap_mode("plot")
tm_shape(co_outline) + tm_polygons(alpha=0) +
tm_shape(sampled_map) + tm_fill("sampled", id="site_name",
palette=c("red","blue","green")) + tm_layout(frame=F)
```

![](map.jpeg)

## Monthly

![](HT_Monthly.jpeg){width=1000px}

## Yearly

![](HT_Obs.jpeg){width=1000px}

## The use of remote sensing data

Landsat data can provide us with valuable data that can show us what the conditions of the reservoirs are at any given time. This data, like the blueness or greenness of a reservoir, can help us look at reservoirs that we do not have data for.

![](ht_color_dist.jpeg){width=1000px}

## Climatology by month

We can see that Horsetooth is more green and has higher wavelengths in the summer months.
This trend is not specific to Horsetooth, as algae blooms will cause this. This trend is why we only care about summer months.

![](HT_Climatology.jpeg){width=1000px}

## What does that mean for the summer months, though?

We can see that over time, during the summer, that Horsetooth is actually becoming more blue. This isn't the case for all reservoirs, as some are actually becoming more green. The greener they are, the more suspected algae is in the water.

![](ht_color_over_time.jpeg){width=1000px}

##

### 2020-12-07 main.py
fix imports

@@ -4,9 +4,7 @@ import numpy as np
 import pandas as pd
 from sklearn.ensemble import GradientBoostingRegressor, RandomForestRegressor
 from sklearn.metrics import mean_squared_error
-from sklearn.model_selection import (cross_val_score,
-                                     RandomizedSearchCV,
-                                     train_test_split)
+from sklearn.model_selection import cross_val_score, train_test_split
 from sklearn.preprocessing import OneHotEncoder
 from xgboost import XGBRegressor
 

### 2020-12-07 plots.py
tidy up notebooks and plots

@@ -1,0 +1,117 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sb
+
+from main import prep_insitu, prep_sat, experiment_dataset, experiment_model
+
+
+# data
+insitu = pd.read_csv('complete_in-situ.csv', low_memory=False)
+sat = pd.read_csv('larger_landsat.csv', low_memory=False)
+insitu = prep_insitu(insitu)
+sat = prep_sat(sat)
+chloro = insitu['chlorophyll']
+log_chloro = np.log(chloro)
+insitu['chlorophyll'] = log_chloro
+insitu = insitu[np.isfinite(log_chloro)]
+
+# exploration
+sb.histplot(chloro, color='xkcd:algae', bins=15)
+sb.despine()
+plt.xlabel('Chlorophyll')
+plt.title('Distribution of Chlorophyll Measurements')
+plt.savefig('figures/chloro_hist.png', dpi=200, bbox_inches='tight')
+
+sb.histplot(log_chloro, color='xkcd:algae', bins=15)
+sb.despine()
+plt.xlabel('Chlorophyll')
+plt.title('Distribution of Chlorophyll Measurements\n(log-transform)')
+plt.savefig('figures/log_chloro_hist.png', dpi=200, bbox_inches='tight')
+
+# feature importance
+model = experiment_model('random_forest')
+variables = ['red', 'green', 'blue', 'nir', 'aerosol', 'areasqkm',
+             'dwl', 'swir1', 'swir2', 'tir1', 'tir2']
+X_train, y_train = experiment_dataset(
+    insitu,
+    sat,
+    subset=variables,
+    merge_thresh=3
+)
+model.fit(X_train, y_train)
+feature_importances = model.feature_importances_
+importance_series = pd.Series(dict(zip(variables, feature_importances)))
+importance_series = importance_series.sort_values(ascending=False)
+sb.barplot(
+    y=importance_series.index,
+    x=importance_series.values,
+    color='xkcd:algae'
+)
+sb.despine()
+plt.title('Feature Importances')
+plt.savefig('figures/feature_importances.png', dpi=200, bbox_inches='tight')
+
+# model performance
+results = pd.read_csv('results.csv')
+current_palette = sb.color_palette()
+palette = {'random_forest': current_palette[0],
+           'gradient_boost': current_palette[1],
+           'xgboost': current_palette[2]}
[…]

### 2020-12-08 paper.Rmd
Add files via upload

---
title: "DSCI 445 - Algal Blooms"
author: "Dawson Eliasen, Boston Lee, Miles Austin"
date: "12/7/2020"
output:
  pdf_document: default
---

  Algae blooms are becoming more relevant in today’s world, due to the increasing 
climate change. Scientific research has shown that algae blooms are happening at
an increasing frequency, with a higher geographic distribution than we had 
previously seen, which is especially notable in high elevation reservoirs. Not 
only are algae blooms occurring more frequently, harmful algal blooms - also 
known as HABS - are also occurring more frequently.

  Even though algae are natural and part of a healthy freshwater ecosystem, the
algae can grow to an extent that it can be harmful to the ecosystem – creating 
anoxic conditions for example, otherwise known as “dead zones” for the aquatic 
fauna. The HABs can produce toxins that can be harmful to not only the aquatic 
ecosystem, but to humans and other animals, such as dogs. Both algae blooms and 
HABs have a wide range of impacts that can not only be local, but they can also
have interstate impacts. This is especially important to note because Lake Erie,
in 2014, had a HAB that was incredibly impactful to the local economy and 
residents. The HAB left half-a-million people without clean water to drink and 
roughly $65 million in economic losses over a two day span of time.

  With the increasing use of reservoirs for municipalities, the quality of the 
water is very important. Although the water gets treated before it gets to the 
tap, the quality of the water as it reaches the water treatment plant is very 
important regarding the cost to treat the water for municipal use, and algal 
blooms impact the quality of the water dramatically. Most municipal water is 
monitored, but not all of the data is publicly available and some isn’t 
monitored in situ - otherwise known as collected by hand. However, Landsat is a
series of satellites that image the Earth surface, in repeated intervals of 14 
days; that is to say Fort Collins gets re-imaged every 14 days. Landsat 5 dates
back to 1984, and is no longer taking images, while landsat 7 (1999) and landsat
8 (2013) still record images today. The landsat imagery allows for a larger 
sample of data to look at the trends of algal blooms in Colorado, in comparison 
to the lack of public in situ data available. Figure 1 shows reservoirs - only 
those greater than 0.01 km² -  that have both in situ and satellite imagery in 
blue, while the reservoirs in red indicate the reservoirs with just the landsat 
imagery available. It is easy to see the lack of reservoir data we have 
available to us, which is why utilizing machine learning to try and predict 
algae blooms from just remotely sensed imagery would be an incredibly useful
method of determining water quality in reservoirs.

Methods:
How we got the data
+/- day merge
Tried 3 models
Tried 5 variables
Tried day threshold
Train/test split? (⅓ was held out)
Experimental variables were compared using 5-fold cross validation

Results:

### 2020-12-08 slides.Rmd
add results slides content and python code

@@
 output:
+  beamer_presentation: default
   slidy_presentation: default
@@
 
-```{r setup, include=FALSE}
-library(tidyverse)
-library(tmap)
-library(tmaptools)
-library(USAboundaries)
-library(sf)
-knitr::opts_chunk$set(echo = FALSE)
+```{r setup, echo=FALSE}
+library(knitr)
+opts_chunk$set(engine.path='venv/bin/python3')
+```
+
+```{python, include=FALSE}
+from plots import *
 ```
@@
 When looking at the amount of hand-collected (in situ) data, we see that in Colorado only has data for the major municipal water bodies like Horsetooth.
-```{r, eval=F}
-sampled_map <- st_read("sampled_map.csv", geometry_column='geometry')
-states <- us_boundaries()
-st_as_sf("sampled_map")
-co_outline <- states %>%
-  filter(name == 'Colorado') %>%
-  st_transform(2163)
-tmap_mode("plot")
-tm_shape(co_outline) + tm_polygons(alpha=0) +
-tm_shape(sampled_map) + tm_fill("sampled", id="site_name",
-palette=c("red","blue","green")) + tm_layout(frame=F)
-```
 
@@
 
-##
+## Methodology
+
+```{python, echo=FALSE}
+plot_histograms()
+```
+
+## Random forest feature importance
+
+- We were surprised to see `areasqkm` as the most important feature
+- `dwl` as the second most important feature makes sense
+
+```{python, echo=FALSE}
+plot_feature_importance()
+```
+
+## How does the feature subset affect performance?
+
+- Different models behave similarly
+- Adding in `areasqkm` resulted in the biggest performance increase
+
+```{python, echo=FALSE}
+plot_subset_performance()
+```
+
+## How does the merge threshold affect performance?
+
+- Different models behave similarly
+- Interestingly, a merge threshold of 1 is worse than 0 and 2
+
+```{python, echo=FALSE}
+plot_thresh_performance()
+```
+
+## Results summary
+
+- Our best model was a random forest with a merge threshold of 3 days and included all available features
+- Our best model received a CV MSE of about 1.75 and a test set MSE of about 1.09
+
+## Predicted values
+
+```{python, echo=FALSE}
+plot_predicted_values()
+```

### 2020-12-08 plots.py

@@ -11,107 +11,150 @@ insitu = pd.read_csv('complete_in-situ.csv', low_memory=False)
 sat = pd.read_csv('larger_landsat.csv', low_memory=False)
 insitu = prep_insitu(insitu)
 sat = prep_sat(sat)
-chloro = insitu['chlorophyll']
+chloro = insitu['chlorophyll'].copy()
 log_chloro = np.log(chloro)
 insitu['chlorophyll'] = log_chloro
 insitu = insitu[np.isfinite(log_chloro)]
-
-# exploration
-sb.histplot(chloro, color='xkcd:algae', bins=15)
-sb.despine()
-plt.xlabel('Chlorophyll')
-plt.title('Distribution of Chlorophyll Measurements')
-plt.savefig('figures/chloro_hist.png', dpi=200, bbox_inches='tight')
-
-sb.histplot(log_chloro, color='xkcd:algae', bins=15)
-sb.despine()
-plt.xlabel('Chlorophyll')
-plt.title('Distribution of Chlorophyll Measurements\n(log-transform)')
-plt.savefig('figures/log_chloro_hist.png', dpi=200, bbox_inches='tight')
-
-# feature importance
-model = experiment_model('random_forest')
-variables = ['red', 'green', 'blue', 'nir', 'aerosol', 'areasqkm',
-             'dwl', 'swir1', 'swir2', 'tir1', 'tir2']
-X_train, y_train = experiment_dataset(
-    insitu,
-    sat,
-    subset=variables,
-    merge_thresh=3
-)
-model.fit(X_train, y_train)
-feature_importances = model.feature_importances_
-importance_series = pd.Series(dict(zip(variables, feature_importances)))
-importance_series = importance_series.sort_values(ascending=False)
-sb.barplot(
-    y=importance_series.index,
-    x=importance_series.values,
-    color='xkcd:algae'
-)
-sb.despine()
-plt.title('Feature Importances')
-plt.savefig('figures/feature_importances.png', dpi=200, bbox_inches='tight')
-
 # model performance
 results = pd.read_csv('results.csv')
+# plot settings
 current_palette = sb.color_palette()
 palette = {'random_forest': current_palette[0],
            'gradient_boost': current_palette[1],
            'xgboost': current_palette[2]}
-# merge threshold vs model performance
-merge_thresh_results = results \
-    .groupby(['model_type', 'merge_thresh'])['mse'] \
-    .min() \
-    .reset_index() \
[…]

### 2020-12-09 paper.Rmd
Added preliminary discussion of methods to both slides and paper.

@@
 ---
+
+```{r setup, echo=FALSE}
+library(knitr)
+opts_chunk$set(engine.path='venv/bin/python3')
+```
+
+# Background
 
@@
 residents. The HAB left half-a-million people without clean water to drink and 
-roughly $65 million in economic losses over a two day span of time.
+roughly \$65 million in economic losses over a two day span of time.
 
@@
 
-Methods:
+# Methods
+
+## Methods TODO:
 How we got the data
+
+## Variable Selection
+
+## Data Manipulation
+
+Tried 5 variables
 +/- day merge
-Tried 3 models
-Tried 5 variables
 Tried day threshold
-Train/test split? (⅓ was held out)
+
+The insitu data was not collected as frequently as the satellite imagery data.
+To remedy this, an experimental approach was taken in which the FIXME data
+was shifted by a certain number of days forward and backward. Intuitively, algae
+blooms do not drastically alter the characteristics of a lake between consecutive days.
+Interpolating surrounding observations with existing data allowed for many more
+of the satellite observations to be used, even though there was no insitu data
+accompanying the observations.
+
+However, the aforementioned Interpolating approach certainly fails after a
+certain range of days. The optimal shift was determined experimentally: Each
+model was fit using data shifted from one to ten days in either direction, and
+the cross-validation error was used to assess model accuracy. The specifics of
+the experimental structure are described below. The model accuracy provided a
+picture of the quality of interpolation provided by each shift. If data was
+shifted too many days, the prediction accuracy of the algorithm would
+deteriorate. FIXME mention optimal shift and include graph.
+
+## Models
+
+The algae bloom data is spatiotemporal; the blooms are recorded in a certain
+geographic location at a certain time. This makes the class of linear models
+unsuitable for this problem. Linear models make assumptions about the structure
+of the errors in the data generating process that are violated by spatiotemporal
+data. Spatiotemporal data has errors which are correlated; as observations
+closer in space or time will covary(CITE).
+
+To work around the issue of violated assumptions in classical statistical
+models, we opted to use three variants of non-parametric, tree based regression
+models. The models used were Random Forest, Gradient Boosting and XGBoost. These
+models do not make assumptions about the structure of errors and correlations in
+the data, meaning spatiotemporal data is usable.
+
+TODO Describe each model?
+
+## Model Selection
+
+Train/test split? ($\frac{1}{3}$ was held out)
 Experimental variables were compared using 5-fold cross validation
 
-Results:
+# Results

### 2020-12-09 slides.Rmd
Added preliminary discussion of methods to both slides and paper.

@@
   beamer_presentation: default
-  slidy_presentation: default
-editor_options:
-  chunk_output_type: console
 ---
@@
 from plots import *
+=======
+```{r setup, include=FALSE}
+library(tidyverse)
+library(tmap)
+library(tmaptools)
+library(USAboundaries)
+library(sf)
+knitr::opts_chunk$set(echo = FALSE)
+>>>>>>> Stashed changes
 ```
@@
 
+## Algorithm Choice
+
+- The algae bloom data is spatiotemporal: We have measurements that are potentially correlated over geographic space, that are taken over time
+- This means we cannot use any regression-based methods. Assumptions about uncorrelated errors are violated.
+- Solution: Non-parametric algorithms
+
+## Algorithm Choice, Cont.
+
+- We chose two non-parametric algorithms:
+  - Random Forest
+  - Gradient Boosting
+
+## Gradient Boosted Trees
+
+- Given a differentiable loss function $L(\cdot)$, we want to **minimize** the total loss for points in the test set:
+
+$$\sum_{i = 1}^{n} L(y_i, F(x_i))$$
+
+Where $F(y_i, F(x_i))$ is a measure of how **wrong** $F(x_i)$ is in predicting $y_i$.
+
+- This leads to an optimization problem, which (because of differentiability) can be solved by gradient descent.
+
+## Gradient Boosted Trees, cont.
+
+- We can interpret residuals as the **negative gradient** of the total loss defined before.
+- Like AdaBoost, we update the prediction function, but now, we update $F(\cdot)$ based on the negative gradient of our total loss.
+- This algorithm can handle different loss functions, with different properties, making it very flexible.
+- Here, we use the default implementation in `scikit-learn`, which has a least-squares loss function:
+
+$$L(y_i, F(x_i)) = (y_i = F(x_i))^2$$
+
 ## Random forest feature importance

### 2020-12-10 paper.Rmd
Added subsections, figures, and captions to paper.

@@
 opts_chunk$set(engine.path='venv/bin/python3')
+```
+
+```{python, include=FALSE}
+from plots import *
 ```
@@
 
-## Methods TODO:
+TODO:
 How we got the data
 
-## Variable Selection
-
-## Data Manipulation
+## Data Interpolation (Merging Shift Threshold)
 
@@
 picture of the quality of interpolation provided by each shift. If data was
-shifted too many days, the prediction accuracy of the algorithm would
-deteriorate. FIXME mention optimal shift and include graph.
+shifted too many days, the prediction accuracy of the algorithm should
+deteriorate.
+
+## Response Variable Transformation
+
+As can be seen from Figure \ref{fig:resp-hist}, the chlorophyll variable, the response
+variable in this case, was initially highly skewed. This had an impact on
+initial prediction accuracy. In order to correct for the skewed nature of the
+data, the response was log-transformed, yielding the second distribution in
+Figure \ref{fig:resp-hist}.
+
+```{python, echo=FALSE, fig.cap="Distribution of chlorophyll before and after log transformation.\\label{fig:resp-hist}"}
+plot_histograms()
+```
+
+It should be noted that all MSE values were calculated from chlorophyll on the
+log-transformed scale, then exponentiated to account for the variable
+transformation.
 
@@
 data. Spatiotemporal data has errors which are correlated; as observations
-closer in space or time will covary(CITE).
+closer in space or time will covary (CITE).
 
@@
 # Results
+
+## Merge Shifting Threshold
+
+Comparing across different shifting thresholds, all three models included in
+the experiment performed similarly.
+
+```{python, echo=FALSE, fig.cap="Cross-validation error for all three models at merge thresholds from 1 to 10.\\label{fig:thresh-performance}"}
+plot_thresh_performance()
+```
+
+As we can see from Figure \ref{fig:thresh-performance}, the trend across merge
+thresholds for each model is very similar. As can also be seen in Figure
+\ref{fig:thresh-performance}, the optimal shifting threshold is three
+days \ref{fig:thresh-performance}. It is also interesting to note that a shift of one day performed
+worse than no shift, or a shift of two days. This is a counterintuitive result,
+but it could be due to the tradeoff between the extra volume of data added by
+larger shift thresholds versus the accuracy of the data at lower shifting
+thresholds.
+
+## Model Performance
+
+All three models mentioned performed similarly, with no model greatly exceeding
+the other two in terms of CV error. The best overall model, in terms of FIXME
+error, was a random forest with a merge shift threshold of three days, that
+included all features. The best model had a CV MSE of approximately 1.75, and a
+test MSE of approximately 1.09.

### 2020-12-10 slides.Rmd

@@
 from plots import *
-=======
-```{r setup, include=FALSE}
-library(tidyverse)
-library(tmap)
-library(tmaptools)
-library(USAboundaries)
-library(sf)
-knitr::opts_chunk$set(echo = FALSE)
->>>>>>> Stashed changes
 ```
@@
 
-When looking at the amount of hand-collected (in situ) data, we see that in Colorado only has data for the major municipal water bodies like Horsetooth.
+When looking at the amount of hand-collected (in situ) data, we see that in
+Colorado only has data for the major municipal water bodies like Horsetooth.
 
@@
 
-![](HT_Monthly.jpeg){width=1000px}
+![](HT_Monthly.jpeg)
 
@@
 
-![](HT_Obs.jpeg){width=1000px}
+![](HT_Obs.jpeg)
 
@@
 
-Landsat data can provide us with valuable data that can show us what the conditions of the reservoirs are at any given time.
-This data, like the blueness or greenness of a reservoir, can help us look at reservoirs that we do not have data for.
+Landsat data can provide us with valuable data that can show us what the
+conditions of the reservoirs are at any given time. This data, like the
+blueness or greenness of a reservoir, can help us look at reservoirs that we do
+not have data for.
 
-![](ht_color_dist.jpeg){width=1000px}
+![](ht_color_dist.jpeg)
 
@@
 
-We can see that Horsetooth is more green and has higher wavelengths in the summer months.
-This trend is not specific to Horsetooth, as algae blooms will cause this. This trend is why we only care about summer months.
+We can see that Horsetooth is more green and has higher wavelengths in the
+summer months. This trend is why we only care about summer months.
 
-![](HT_Climatology.jpeg){width=1000px}
+![](HT_Climatology.jpeg)
 
-## What does that mean for the summer months, though?
+## Summer Months
 
-We can see that over time, during the summer, that Horsetooth is actually becoming more blue.
-This isn't the case for all reservoirs, as some are actually becoming more green.
-The greener they are, the more suspected algae is in the water.
+We can see that over time, during the summer, that Horsetooth is actually
+becoming more blue. This isn't the case for all reservoirs, as some are
+actually becoming more green. The greener they are, the more suspected algae is
+in the water.
 
-![](ht_color_over_time.jpeg){width=1000px}
+![](ht_color_over_time.jpeg)
+
+## Summer Months
+
+- Over time, during the summer, Horsetooth is becoming more blue. 
+- This is not the case for all reservoirs; some are actually becoming more
+  green. 
+- The greener they are, the more suspected algae is in the water.
+
+![](ht_color_over_time.jpeg)
 
@@
 
-- The algae bloom data is spatiotemporal: We have measurements that are potentially correlated over geographic space, that are taken over time
-- This means we cannot use any regression-based methods. Assumptions about uncorrelated errors are violated.
+- The algae bloom data is spatiotemporal: We have measurements that are
+  potentially correlated over geographic space, that are taken over time
+- This means we cannot use any regression-based methods. Assumptions about
+  uncorrelated errors are violated.
 - Solution: Non-parametric algorithms
@@
 
-- We can interpret residuals as the **negative gradient** of the total loss defined before.
-- Like AdaBoost, we update the prediction function, but now, we update $F(\cdot)$ based on the negative gradient of our total loss.
-- This algorithm can handle different loss functions, with different properties, making it very flexible.
-- Here, we use the default implementation in `scikit-learn`, which has a least-squares loss function:
+- We can interpret residuals as the **negative gradient** of the total loss
+  defined before.
+- Update $F(\cdot)$ based on the negative gradient of our total loss.
+- This algorithm can handle different loss functions, with different
+  properties, making it very flexible.
+- Here, we use the default implementation in `scikit-learn`, which has a
+  least-squares loss function:
 

### 2020-12-10 paper.Rmd
Added how we got the data

@@
 
-TODO:
-How we got the data
+The data came from two sources; satellite data and in situ data from public
+sources. The satellite data was downloaded utilizing Google Earth Engine, which
+downloaded the data for individual lakes, ignoring images that were over a
+threshold of cloud cover. This threshold was because cloud cover would darken
+the pixels, skewing the data. The satellite data included all of Landsat 5, 7,
+and 8 from the past 35 year and has the different band (wavelength) data for the
+center most pixel of each lake. The in situ data includes all public data that
+is available for chlorophyll in Colorado. This includes data from sources like
+Denver Water, the USGS, and Northern Water. The data that is being utilized is
+chlorophyll values, and the data also includes secchi depth (how deep you can
+see into the water).
 

### 2020-12-10 slides.Rmd
Update slides.Rmd

@@
 author: "Dawson Eliasen, Boston Lee, Miles Austin"
-date: "12/7/2020"
+date: "12/11/2020"
 output:
@@
 
-## Monthly
+## Where did the data come from?
+
+We are utilizing satellite data that was gathered from GEE & all in situ public
+water quality data for Colorado.
+  - The satellite images are collected every 14 days.
+  - In situ data is collected randomly.
+
+## A look at the monthly data
 
@@
 
-## Yearly
+## A look at the yearly data
 
@@
 ![](HT_Climatology.jpeg)
-
-## Summer Months
-
-We can see that over time, during the summer, that Horsetooth is actually
-becoming more blue. This isn't the case for all reservoirs, as some are
-actually becoming more green. The greener they are, the more suspected algae is
-in the water.
-
-![](ht_color_over_time.jpeg)
 
@@
 
-$$L(y_i, F(x_i)) = (y_i = F(x_i))^2$$
+$$L(y_i, F(x_i)) = (y_i - F(x_i))^2$$
 

### 2020-12-10 paper.Rmd
Added in the where we got the data

@@
 important regarding the cost to treat the water for municipal use, and algal 
-blooms impact the quality of the water dramatically. Most municipal water is 
+blooms impact the quality of the water dramatically(Brookes). Most municipal water is 
 monitored, but not all of the data is publicly available and some isn’t 
@@
 algae blooms from just remotely sensed imagery would be an incredibly useful
-method of determining water quality in reservoirs.
+method of determining water quality in reservoirs. It has been done previously,
+but mainly utilizing only the blue and green wavelengths, despite there being more
+potential options(Sachidananda).
 
@@
 test MSE of approximately 1.09.
+
+## Sources
+
+Brezonik, Patrick, Kevin D. Menken, and Marvin Bauer. “Landsat-Based Remote
+Sensing of Lake Water Quality Characteristics, Including Chlorophyll and Colored
+Dissolved Organic Matter (CDOM).” Lake and Reservoir Management 21, no. 4 
+(December 2005): 373–82. https://doi.org/10.1080/07438140509354442.
+
+Brookes, J. D., and C. C. Carey. “Resilience to Blooms.” Science 334, no. 6052 
+(October 7, 2011): 46–47. https://doi.org/10.1126/science.1207349.
+
+Brookes, Justin D., Cayelan C. Carey, David P. Hamilton, Lionel Ho, Leon van der
+Linden, Robert Renner, and Anna Rigosi. “Emerging Challenges for the Drinking 
+Water Industry.” Environmental Science & Technology 48, no. 4 (February 18, 
+2014): 2099–2101. https://doi.org/10.1021/es405606t.
+
+Folke, Carl, Steve Carpenter, Brian Walker, Marten Scheffer, Thomas Elmqvist,
+Lance Gunderson, and C.S. Holling. “Regime Shifts, Resilience, and Biodiversity 
+in Ecosystem Management.” Annual Review of Ecology, Evolution, and Systematics 
+35, no. 1 (December 15, 2004): 557–81. 
+https://doi.org/10.1146/annurev.ecolsys.35.021103.105711.
+
+Ho, Jeff C., Anna M. Michalak, and Nima Pahlevan. “Widespread Global Increase in
+Intense Lake Phytoplankton Blooms since the 1980s.” Nature 574, no. 7780 
+(October 2019): 667–70. https://doi.org/10.1038/s41586-019-1648-7.
+
+Olmanson, Leif G, Marvin E Bauer, and Patrick L Brezonik. “Use of Landsat
+Imagery to Develop a Water Quality Atlas of Minnesota’s 10,000 Lakes.” 
+Conference Proceedings, 2002, 8.
+
+Scheffer, Marten, Steve Carpenter, Jonathan A. Foley, Carl Folke, and Brian 
+Walker. “Catastrophic Shifts in Ecosystems.” Nature 413, no. 6856 (October 2001)
+591–96. https://doi.org/10.1038/35098000.
+
+Torbick, Nathan, Sarah Hession, Stephen Hagen, Narumon Wiangwang, Brian Becker, 
+and Jiaguo Qi. “Mapping Inland Lake Water Quality across the Lower Peninsula of
+Michigan Using Landsat TM Imagery.” International Journal of Remote Sensing 34,
+no. 21 (November 10, 2013): 7607–24. https://doi.org/10.1080/01431161.2013.822602.

### 2020-12-11 slides.Rmd
Update slides.Rmd
fix image paths
Added suggested changes.

@@
 
-![](map.jpeg)
+![](images/map.jpeg)
 
@@
 
-![](HT_Monthly.jpeg)
+![](images/HT_Monthly.jpeg)
 
@@
 
-![](HT_Obs.jpeg)
+![](images/HT_Obs.jpeg)
 
@@
 
-![](ht_color_dist.jpeg)
+![](images/ht_color_dist.jpeg)
 
@@
 
-![](HT_Climatology.jpeg)
+![](images/HT_Climatology.jpeg)
 
@@
 
-![](ht_color_over_time.jpeg)
+![](images/ht_color_over_time.jpeg)
 
-## Methodology
+## Merge Threshold
+
+- In-situ and satellite data do not necessarily have matching dates
+- On an inner join, most of the data would be dropped
+- Solution: Shift in-situ data $0, \cdots, n$ days forward and backward
+
+## Response Variable Transformation
+
+Because of extreme skew, the response variable, chlorophyll, was
+log-transformed.
 
@@
 ```
+
+## Model selection
+
+- Final model was selected via 5-fold cross-validation error on a training set.
+- Models were fit over a grid of:
+  1. Model type
+  1. Variable Subset
+  1. Merge Threshold
 
@@
 
+![](images/gradient_boosting_simple.png)
+
+Image credit: https://arogozhnikov.github.io/2016/06/24/gradient_boosting_explained.html
+
 ## Gradient Boosted Trees
@@
 
-- We can interpret residuals as the **negative gradient** of the total loss
-  defined before.
-- Update $F(\cdot)$ based on the negative gradient of our total loss.
-- This algorithm can handle different loss functions, with different
-  properties, making it very flexible.
+- We can interpret residuals as the **negative gradient** of the total loss.
 - Here, we use the default implementation in `scikit-learn`, which has a

### 2020-12-14 paper.Rmd

@@
 title: "DSCI 445 - Algal Blooms"
-author: "Dawson Eliasen, Boston Lee, Miles Austin"
-date: "12/7/2020"
+author: "Miles Austin, Dawson Eliasen, Boston Lee"
+date: "12/14/2020"
 output:
@@
 
-# Background
+# Motivation
 
@@
 
+![](images/map.jpeg)
+
+![](images/HT_Climatology.jpeg)
+
+![](images/ht_color_over_time.jpeg)
+
 # Methods
@@
 
-The insitu data was not collected as frequently as the satellite imagery data.
+The in situ data was not collected as frequently as the satellite imagery data.
 To remedy this, an experimental approach was taken in which the FIXME data
@@
 Interpolating surrounding observations with existing data allowed for many more
-of the satellite observations to be used, even though there was no insitu data
+of the satellite observations to be used, even though there was no in situ data
 accompanying the observations.
@@
 
+Overall, random forest, gradient boost, and XGBoost performed similarly, with random forest
+performing slightly better than the other two model types.
+
+## Feature subsets and importance
+
+Performance for all three models generally improved as more features were added.
+The inclusion of the lake area feature resulted in the largest performance improvement.
+\ref{fig:feature-importance} shows the feature importance statistics for a random forest
+model fit to a dataset constructed with a merge threshold of 3 days and including all features.
+
+```{python, echo=FALSE, fig.cap="Feature importance from random forest with all features included.\\label{fig:feature-importance}"}
+plot_feature_importance()
+```
+
 ## Merge Shifting Threshold
 
-Comparing across different shifting thresholds, all three models included in
-the experiment performed similarly.
+When comparing results across different merge threshold values, we see all three model
+types perform similarly.
 
@@
 As we can see from Figure \ref{fig:thresh-performance}, the trend across merge
-thresholds for each model is very similar. As can also be seen in Figure
-\ref{fig:thresh-performance}, the optimal shifting threshold is three
-days \ref{fig:thresh-performance}. It is also interesting to note that a shift of one day performed
-worse than no shift, or a shift of two days. This is a counterintuitive result,
+thresholds for each model is very similar. Generally, performance decreases before increasing
+and staying relatively high. It interesting to note that a shift of one day performed worse 
+than no shift, or a shift of two days. This is a counter-intuitive result,
 but it could be due to the tradeoff between the extra volume of data added by
@@
 All three models mentioned performed similarly, with no model greatly exceeding
-the other two in terms of CV error. The best overall model, in terms of FIXME
+the other two in terms of CV error. The best overall model, in terms of CV MSE
 error, was a random forest with a merge shift threshold of three days, that
 included all features. The best model had a CV MSE of approximately 1.75, and a
-test MSE of approximately 1.09.
+test set MSE of approximately 1.09.
+
+## Predicted Values
+
+Finally, we generate chlorophyll estimates for each of the available satellite images,
+including satellite images of lakes for which there is no available in situ data.
+\ref{fig:predicted-values} Shows the estimated chlorophyll value against the dominant
+wavelength of the satellite image. We can see that the highest chlorophyll estimates
+are attributed to satellite images which have a green dominant wavelength. However, we
+are generating predictions of chlorophyll values exceeding $70 \frac{\mu g}{mL}$, which
+is most likely unreasonable. This is likely an artifact of the extreme observations included
+in the training data with chlorophyll measurements exceeding $300 \frac{\mu g}{mL}$.
+
+```{python, echo=FALSE, fig.cap="Predicted values vs. dominant wavelength.\\label{fig:predicted-values}"}
+plot_predicted_values()
+```
+
+## Conclusion
+
+We have described a process for using non-parametric machine learning methods to estimate
+chlorophyll concentrations in lakes from satellite images. This process broadly consists of
+these steps.
+
+1. Collect satellite image data.
+2. Process satellite image data to create features.
+3. Associate image features with in situ measurements, allowing some small number of days in error.
+4. $\log$-transform the chlorophyll measurements to unskew the distribution.
+5. Fit a model to these data, with $\log$-transform chlorophyll measurements as the response variable and satellite features as the predictor variables.
+6. Use this model to generate predicted $\log (\text{chlorophyll})$ values for each satellite image.
+7. Exponentiate the predicted values to return them to original scale.
+
+Using this process, we investigated the effectiveness of different model types, feature subsets,
+and merge threshold values and found the best results with a random forest model using all features
+and a merge threshold value equal to three days. This model had a 5-fold cross validation MSE of 
+1.75, and the MSE on the holdout test set was 1.05. Based on these error measurements,
+we have demonstrated that these models are capable of estimating chlorophyll concentrations
+with practical accuracy. This level of accuracy is likely at least partially attributable to the fact
+that the training data includes repeated sampling of a handful of lakes, and as a result the error
+measurements are measuring the model's ability to estimate chlorophyll concentrations from *unseen*
+images of *familiar* lakes. To evaluate the predictive accuracy of these models on *unfamiliar* lakes,
+an analyst would require a dataset observing a much larger number and range of lakes. As mentioned
+previously, only measurements from municipal water sources were available.
+
+In future exploration, we hope to address the extreme predicted values by removing observations with a
+chlorophyll concentration exceeding $15 \frac{\mu g}{mL}$. We expect this change to result in a much
+more reasonable range of predicted values. We also think that, based on the importance of the lake area
+feature, the inclusion of a feature describing lake elevation will be important
 

### 2020-12-14 paper.Rmd
Added changes to explanation of models and merge threshold.

@@
 
-Tried 5 variables
-+/- day merge
-Tried day threshold
+Satellite image data and in situ data were not collected on similar schedules.
+This meant that many observations were lost from both datasets on an inner join.
+To remedy this problem, a procedure was used referred to as a "merge threshold."
+Using a merge threshold, the in situ data was shifted forward and backward $0,
+1, \cdots, n$ days. For example, on a merge threshold of two, the in situ data
+from any single observation on day $k$ would be used on day $k - 2, k - 1, k, k
++ 1$, and $k + 2$. In other words, the data from any single in situ observation
+  would be used for the surrounding four days. The merge threshold procedure was
+  done up to ten days of threshold.
 
-The in situ data was not collected as frequently as the satellite imagery data.
-To remedy this, an experimental approach was taken in which the FIXME data
-was shifted by a certain number of days forward and backward. Intuitively, algae
-blooms do not drastically alter the characteristics of a lake between consecutive days.
-Interpolating surrounding observations with existing data allowed for many more
-of the satellite observations to be used, even though there was no in situ data
-accompanying the observations.
-
-However, the aforementioned Interpolating approach certainly fails after a
+However, the aforementioned approach certainly fails after a
 certain range of days. The optimal shift was determined experimentally: Each
@@
 data. Spatiotemporal data has errors which are correlated; as observations
-closer in space or time will covary (CITE).
+closer in space or time will covary.
 
-To work around the issue of violated assumptions in classical statistical
-models, we opted to use three variants of non-parametric, tree based regression
-models. The models used were Random Forest, Gradient Boosting and XGBoost. These
-models do not make assumptions about the structure of errors and correlations in
-the data, meaning spatiotemporal data is usable.
+The models selected were tree-based: Random Forest, Boosted Trees, and XGBoost. Although
+the classical formulation of Random Forests and Boosted Trees rely on the
+assumption of independent and identically distributed observations, tree-based
+models were chosen in hopes that the extra flexibility in these models would
+help account for the temporal structure in the data. Unfortunately, there was no
+explicit spatial variable that could be included in the data to account for
+spatial dependence.
 
-TODO Describe each model?
+The Random Forest and Bosted Tree implementation from `scikit-learn` were used.
+The XGBoost implementation was taken from the Python package `xgboost`.
 
@@
 
-Train/test split? ($\frac{1}{3}$ was held out)
-Experimental variables were compared using 5-fold cross validation
+The chosen models were fit on a grid of parameters. The parameters altered were
+merge threshold, variable subset, and type of model. That is, we fit every
+possible combination of merge threshold from $0, \cdots, 10$, the selected
+variable subsets, and the selected model types.
+
+Within each of the aforementioned combinations, a training set and test set were
+created from the merged data, with the training set compromising two-thirds of
+the dataset, and the test set compromising one-third of the dataset.
+
+After a training and test set were created, $5$-fold cross-validation was
+performed. Models were compared via $5$-fold cross validation error on the
+training dataset. The final model was selected via this error, and then results
+were repoted on the test data set.
 

### 2020-12-14 paper.Rmd
cleanup

@@
 monitored, but not all of the data is publicly available and some isn’t 
-monitored in situ - otherwise known as collected by hand. However, Landsat is a
-series of satellites that image the Earth surface, in repeated intervals of 14 
-days; that is to say Fort Collins gets re-imaged every 14 days. Landsat 5 dates
-back to 1984, and is no longer taking images, while landsat 7 (1999) and landsat
-8 (2013) still record images today. The landsat imagery allows for a larger 
-sample of data to look at the trends of algal blooms in Colorado, in comparison 
-to the lack of public in situ data available. Figure 1 shows reservoirs - only 
-those greater than 0.01 km² -  that have both in situ and satellite imagery in 
-blue, while the reservoirs in red indicate the reservoirs with just the landsat 
-imagery available. It is easy to see the lack of reservoir data we have 
-available to us, which is why utilizing machine learning to try and predict 
-algae blooms from just remotely sensed imagery would be an incredibly useful
-method of determining water quality in reservoirs. It has been done previously,
-but mainly utilizing only the blue and green wavelengths, despite there being more
-potential options(Sachidananda).
+monitored in situ - otherwise known as collected by hand.
 
-![](images/map.jpeg)
-
-![](images/HT_Climatology.jpeg)
-
-![](images/ht_color_over_time.jpeg)
-
-# Methods
-
-The data came from two sources; satellite data and in situ data from public
-sources. The satellite data was downloaded utilizing Google Earth Engine, which
+We will use two distinct datasets: satellite data and in situ data from public
+sources. The satellite data were downloaded from Google Earth Engine, which
 downloaded the data for individual lakes, ignoring images that were over a
@@
 Denver Water, the USGS, and Northern Water. The data that is being utilized is
-chlorophyll values, and the data also includes secchi depth (how deep you can
-see into the water).
+chlorophyll values, which act as a proxy for algae blooms.
+
+![Observed and unobserved lakes in Colorado. \label{map}](images/map.jpeg)
+
+Landsat is a series of satellites that image the Earth surface, in repeated 
+intervals of 14  days; that is to say an individual location such as Fort Collins gets
+re-imaged every 14 days. Landsat 5 dates back to 1984, and is no longer taking images,
+while landsat 7 (1999) and landsat 8 (2013) still record images today. The landsat
+imagery allows for a larger sample of data to look at the trends of algal blooms in
+Colorado, in comparison to the lack of public in situ data available. \ref{fig:map}
+shows reservoirs - only those greater than 0.01 km² -  that have both in situ and
+satellite imagery in blue, while the reservoirs in red indicate the reservoirs with 
+just the landsat imagery available. It is easy to see the lack of reservoir data we have 
+available to us, which is why utilizing machine learning to try and predict 
+algae blooms from just remotely sensed imagery would be an incredibly useful
+method of determining water quality in reservoirs. It has been done previously,
+but mainly utilizing only the blue and green wavelengths, despite there being more
+potential options(Sachidananda).
+
+![Color (dominant wavelength) distributions for Horsetooth Reservoir in each month. \label{climatology}](images/HT_Climatology.jpeg){width=75%}
+
+![Color (dominant wavelength) of Horsetooth Reservoir over time. \label{color-trend}](images/ht_color_over_time.jpeg){width=75%}
+
+We attempted to detect algae blooms from satellite images by using statistical learning
+methods to approximate the relationship between features in satellite images and in situ
+chlorophyll measurements from various lakes in Colorado (predominantly municipal water
+sources such as Horsetooth Reservoir). We reduced the satellite images to a set of
+key features that are likely to be associated with algae blooms; red, green, and blue
+wavelength range measurements, dominant wavelength, and various infrared range measurements.
+We joined these data to in situ chlorophyll measurements and fit various non-parametric
+models before finally producing chlorophyll estimates for all available satellite images.
+
+# Methods
 
@@
 To remedy this problem, a procedure was used referred to as a "merge threshold."
-Using a merge threshold, the in situ data was shifted forward and backward $0,
+Using a merge threshold, the satellite data were shifted forward and backward $0,
 1, \cdots, n$ days. For example, on a merge threshold of two, the in situ data
@@
 + 1$, and $k + 2$. In other words, the data from any single in situ observation
-  would be used for the surrounding four days. The merge threshold procedure was
-  done up to ten days of threshold.
+would be used for the surrounding four days, but each in situ observation would
+only be included once. The merge threshold procedure was done up to ten days of
+threshold.
 
@@
 of the errors in the data generating process that are violated by spatiotemporal
-data. Spatiotemporal data has errors which are correlated; as observations
-closer in space or time will covary.
+data. Spatiotemporal data have errors which are correlated; as observations
+closer in space or time will vary together.
 
@@
 models were chosen in hopes that the extra flexibility in these models would
-help account for the temporal structure in the data. Unfortunately, there was no
-explicit spatial variable that could be included in the data to account for
-spatial dependence.
+help account for the temporal structure in the data.
 
-The Random Forest and Bosted Tree implementation from `scikit-learn` were used.
+The Random Forest and Boosted Tree implementation from `scikit-learn` were used.
 The XGBoost implementation was taken from the Python package `xgboost`.
@@
 training dataset. The final model was selected via this error, and then results
-were repoted on the test data set.
+were reported on the test data set.
 
@@
 
-Finally, we generate chlorophyll estimates for each of the available satellite images,
+Finally, we generated chlorophyll estimates for each of the available satellite images,
 including satellite images of lakes for which there is no available in situ data.
@@
 more reasonable range of predicted values. We also think that, based on the importance of the lake area
-feature, the inclusion of a feature describing lake elevation will be important
+feature, the inclusion of a feature describing lake elevation will be important. Lastly, although we
+included indicators for month to account for variance in time, we wish to repeat the analysis including
+spatial features such as latitude and longitude to account for variance in space.
 
